---
output:
  pdf_document: default
  html_document: default
---


```{r date, echo=FALSE}

date_today <- Sys.Date()
```

---
title: "Statistical Inference Project-Part 2"
author: "Shovit Bhari"
date: `r date_today`
output: html_document
---


# Introduction


This project aanlyzes the "ToothGrowth" dataset which is included in R. The effect of Vitamin C in the length od odontoblasts(cells responsible for tooth growth)  is observed in 60 guinea pigs.Each animal received one of three dose level of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice(OJ) or ascorbic acid (VC). The measurement is done on 10 guinea pigs. 
 [Summary](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html)


## Load Data
Load the libraries required
```{r setup, message=FALSE, warning=FALSE, echo=TRUE}
library(ggplot2)
library(gridExtra)
library(knitr)
library(datasets)
knitr::opts_chunk$set(echo = TRUE, results = 'hold')
```

Load the data and check the summary.
```{r}
data(ToothGrowth)
```

```{r}
summary(ToothGrowth)
```

```{r}
head(ToothGrowth)
```

```{r}
# split of cases between different dose levels and delivery methods
table(ToothGrowth$dose, ToothGrowth$supp)
```


#```{r, echo=FALSE}
#by(ToothGrowth$len, INDICES = list(ToothGrowth$supp, ToothGrowth$dose), length)
#```

```{r}
#Summary of data within each combination of dose level and delivery method:
by(ToothGrowth$len, INDICES = list(ToothGrowth$supp, ToothGrowth$dose), summary)
```
```{r}
unique(ToothGrowth$supp)

```
```{r}
unique(ToothGrowth$dose)
```

```{r}
unique(ToothGrowth$len)

```

Scatterplot comparision between tooth length and Vitamin C dose:

```{r}
g1<- ggplot(aes(x=dose, y = len), data = ToothGrowth) + 
        geom_point(aes(color = supp)) +
        ggtitle("Dispersion of tooth growth by dose") +
        theme(plot.title = element_text(hjust=0.5)) +
        xlab("dose in mg") +
        ylab("tooth length")
g2<- ggplot(aes(x=supp, y = len), data = ToothGrowth) + 
        geom_point(aes(color = dose)) +
        ggtitle("Dispersion of tooth growth by delivery method") +
        theme(plot.title = element_text(hjust=0.5)) +
        xlab("supp") +
        ylab("tooth length")
grid.arrange(g1, g2, nrow=2)
```

```{r}
#Boxplot comparision between Vitamin C dose level and tooth length:
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
g3<- ggplot(ToothGrowth, aes(x=dose, y=len)) + 
        geom_boxplot(aes(fill=dose)) + facet_grid(~supp) +
        ggtitle("Tooth Length vs. Dose Amount \nby Delivery Method") +
        theme(plot.title = element_text(hjust=0.5)) +
        xlab("Dose Amount (mg)") +
        ylab("Tooth Length")
#Boxplot comparision between delivery methods at each dose level:
g4<-ggplot(aes(x = supp, y = len), data = ToothGrowth) +
        geom_boxplot(aes(fill = supp)) + facet_wrap(~ dose) +
        ggtitle("Tooth Length vs. Delivery Method by Dose Amount") +
        theme(plot.title = element_text(hjust=0.5)) +
        xlab("Supplement Delivery") +
        ylab("Tooth Length")
grid.arrange(g3, g4, ncol=2)
```




Comparing tooth growth by supplement using t-test. 
```{r dist, message=FALSE}
t.test(len ~ supp, paired = F, var.equal = F, data = ToothGrowth)
```

Comparing tooth growth by dose, looking at the different pairs do dose values
```{r}
#dose amounts 0.5 and 1.0
dose1 <- subset(ToothGrowth, dose %in% c(0.5, 1.0))
t.test(len ~ dose, paired = F, var.equal = F, data = dose1)

```
```{r}
#dose amounts 0.5 and 2.0
dose2 <- subset(ToothGrowth, dose %in% c(0.5, 2.0))
t.test(len ~ dose, paired = F, var.equal = F, data = dose2)

```
```{r}
#dose amounts 1.0 and 2.0
dose3 <- subset(ToothGrowth, dose %in% c(1.0, 2.0))
t.test(len ~ dose, paired = F, var.equal = F, data = dose3)
```
