
# Exploratory Data Analysis

## Unigram Analysis

Plotting the top ten Unigram frequency.
```{r}

topUniVector <- topfeatures(unigram, 10)
topUniVector <- sort(topUniVector, decreasing = FALSE)
topUniDf <- data.frame(words = names(topUniVector), freq = topUniVector)
topUniPlot <- ggplot(data = topUniDf, aes(x = reorder(words, freq), y = freq, fill = freq)) + 
        geom_bar(stat = "identity") +
        theme_minimal() +
        labs(x = "Unigram", y = "Frequency") +
        labs(title = "Ten most common Unigrams") +
        theme(plot.title = element_text(hjust = 0.5))+ 
        coord_flip() +
        guides(fill=FALSE) 
g1<- ggplotly(topUniPlot)
g1
```

## Bigram Analysis
Plotting the top ten Bigram frequency.
```{r}

topBiVector <- topfeatures(bigram, 10)
topBiVector <- sort(topBiVector, decreasing = FALSE)
topBiDf <- data.frame(words = names(topBiVector), freq = topBiVector)
topBiPlot <- ggplot(data = topBiDf, aes(x = reorder(words, freq), y = freq, fill = freq)) + 
        geom_bar(stat = "identity") +
        theme_minimal() +
        labs(x = "Bigram", y = "Frequency") +
        labs(title = "Ten most common Bigrams") +
        theme(plot.title = element_text(hjust = 0.5)) +
        coord_flip() +
        guides(fill=FALSE) 
g2<- ggplotly(topBiPlot)
g2
```

## Trigram Analysis

Plotting the top ten Trigram frequency.
```{r}

topTriVector <- topfeatures(trigram, 10)
topTriVector <- sort(topTriVector, decreasing = FALSE)
topTriDf <- data.frame(words = names(topTriVector), freq = topTriVector)
topTriPlot <- ggplot(data = topTriDf, aes(x = reorder(words, freq), y = freq, fill = freq)) + 
        geom_bar(stat = "identity") +
        theme_minimal() +
        labs(x = "Trigram", y = "Frequency") +
        labs(title = "Ten most common Trigrams") +
        theme(plot.title = element_text(hjust = 0.5)) +
        coord_flip() +
        guides(fill=FALSE) 
g3<- ggplotly(topTriPlot)
g3
```


# Plans for prediction algorithm and Shiny App

The final part of this capstone project is to build a predictive algorithm that will be deployed as a Shiny app for the user interface in the web browser. In Shiny app will take multiple words input in a text box and output a prediction of the next word. 

The predictive algorithm will be created using an n-gram model with a frequency lookup to that performed in the exploratory data analysis section of this report.Based on the exploratory analysis here are the plans:

* Increase the number of n-grams, possibly upto 5 or 6.  
* Create a model, first to look for the unigram from the entered text.   
* Once the text is entered, find the most common bigram model and increase the n-grams.   
* Increase or decrease the percentage of training data for efficiency and accuracy. 
* Efficiency and accuracy will be a deciding factor or the final strategy.    
* The design for the Shiny App is to be decided. 
